---
title: "Processing File: Personality and race IAT-study"
output: html_document
date: "2023-11-27"
---
```{r, include=FALSE}

knitr::opts_chunk$set(message = FALSE,
                      warning = FALSE)

# disable scientific notation
options(scipen = 999) 

```

# Dependencies

```{r}
library(dplyr)
library(tidyr)
library(janitor)
library(stringr)
library(readr)
library(openxlsx)
```

# Get data

```{r}
data_raw_bfi <- read_csv("../data/raw/data_raw_bfi.csv") |>
  janitor::clean_names() 

data_raw_iat <- read_csv("../data/raw/data_raw_iat.csv", skip = 1) |>
  janitor::clean_names()

data_raw_demographics <- read_csv("../data/raw/data_raw_demographics.csv") |>
  janitor::clean_names()
```

# Demographics

Extract age and gender from the raw demographics data.

```{r}
#remove rows with missing unique_id
data_demographics_unique_id <- data_raw_demographics |> 
  filter(!is.na(unique_id))

#transform demographics into wide format and check if sex has only "f" and "m" as answers and if age is a number
dat_age_gender <- data_demographics_unique_id |>
  pivot_wider(names_from = variable,
              values_from = response) |>
  mutate(sex = case_when(sex %in% c("f", "m") ~ sex,
                         TRUE ~ "missing/other/error"),
         age = case_when(str_detect(age, "^[0-9]+$") ~ age, 
                         TRUE ~ "other/missing/error"))
```

# Exclusions / data quality

## BFI

### Reverse scores the negatively worded items

Reverse score the negatively worded items: the extroversion scale items 2, 5 and 7, conscientiousness items 2, 4 5 and 9, neuroticism items 2, 5, and 7, agreeableness 1, 3, 6, and 8, and openness items 7 and 9.

```{r}
# List of columns to reverse score for each personality trait
personality_traits <- list(
  extroversion = c("bfi_e2", "bfi_e5", "bfi_e7"),
  conscientiousness = c("bfi_c2", "bfi_c4", "bfi_c5", "bfi_c9"),
  neuroticism = c("bfi_n2", "bfi_n5", "bfi_n7"),
  agreeableness = c("bfi_a1", "bfi_a3", "bfi_a6", "bfi_a8"),
  openness = c("bfi_o7", "bfi_o9")
)

# Function to reverse score
reverse_score <- function(data, cols) {
  data %>%
    mutate(across(all_of(cols), ~7 - .))
}

# Apply reverse scoring
data_raw_bfi <- data_raw_bfi

for (trait in personality_traits) {
  data_raw_bfi <- reverse_score(data_raw_bfi, trait)
}

```

### Sanity checks for Item reversal

Include a sanity check that assesses whether these list of item reversals, and your implementation of them, is likely to be correct: For each subscale, create a correlation table among the items (after reversals) and check that all correlations are positive. In general, negative correlations among items are one indication that information about item reversals, or their implementation, is not correct.  

```{r}
# Function to perform correlation check
check_correlations <- function(data, trait) {
  relevant_cols <- grep(paste0("bfi_[", trait, "]\\d"), names(data), value = TRUE)
  
  cor_matrix <- cor(data[relevant_cols], use = "complete.obs")
  return(cor_matrix)
}

# List of personality traits
traits <- c("a", "e", "c", "o", "n")

# Correlation
correlation_results <- lapply(traits, function(trait) {
  i <- paste0("bfi_", trait)
  cor_matrix <- check_correlations(data_raw_bfi, trait)
  
  print(cor_matrix)
  
  # Check if all correlations are positive
  message <- ifelse(all(cor_matrix > 0),
                    paste0("All the correlations of the subscale ", i, " of the BFI are positive."),
                    paste0("NOT all of the correlations of the subscale ", i, " of the BFI are positive."))
  
  print(message)
  
  return(cor_matrix)
})

names(correlation_results) <- paste0("bfi_", traits)

```

### Logical checks of minimum and maximum 

Check that the item level data does not violate the logical minimum and maximum scores (1 to 6). Create an exclusion variable and set participants with impossible data to "exclude". 

```{r}
data_bfi_completeness <- data_raw_bfi |>
  rowwise() |>
  mutate(exclude_impossible_data = ifelse(any(c_across(starts_with("bfi_")) < 1 | c_across(starts_with("bfi_")) > 6, na.rm = TRUE), "exclude", "include")) 
```

### Check for complete data 

Check that all participants have complete data on the BFI scales they completed. Create an exclusion variable and set participants with incomplete data to "exclude".

```{r}
data_bfi_completeness <- data_bfi_completeness %>%
  mutate(
    exclude_incomplete_data = ifelse(
      all(complete.cases(across(starts_with(c("bfi_a", "bfi_o"))))) ||
      all(complete.cases(across(starts_with(c("bfi_c", "bfi_e", "bfi_n"))))),
      "include",
      "exclude"
    )
  ) 
```

### Mean-score subscales

Mean-score the subscales of the BFI scale. Each participant only got either 2 or 3 subscales. 

```{r}
data_bfi_completeness$mean_bfi_a <- rowMeans(select(data_bfi_completeness,starts_with("bfi_a")),
                                             na.rm = TRUE)
data_bfi_completeness$mean_bfi_o <- rowMeans(select(data_bfi_completeness,starts_with("bfi_o")),
                                             na.rm = TRUE)
data_bfi_completeness$mean_bfi_c <- rowMeans(select(data_bfi_completeness,starts_with("bfi_c")),
                                             na.rm = TRUE)
data_bfi_completeness$mean_bfi_e <- rowMeans(select(data_bfi_completeness,starts_with("bfi_e")),
                                             na.rm = TRUE)
data_bfi_completeness$mean_bfi_n <- rowMeans(select(data_bfi_completeness,starts_with("bfi_n")),
                                             na.rm = TRUE)
```

### Check violations mean score

Check that the mean scores do not violate the min and max possible score (i.e., first determine this min and max score), and revise your scoring code if it does. 

```{r}
# min and max possible score is 1 and 7
out_of_range_counts <- data_bfi_completeness %>%
  rowwise() %>%
  summarize(
    across(starts_with("mean_bfi"), ~ sum(!is.na(.) && (. > 7 || . < 1), na.rm = TRUE))
  )
# Print the result
if (any(out_of_range_counts > 0)) {
  print("Some mean scores violate the min and max possible score (1 and 7). Please revise your scoring code.")
} else {
  print("All mean scores are within the valid range.")
}
```

## IAT

### Scoring the trial-level IAT data

Score the trial-level IAT data using the Greenwald "D" score: Calculate a mean RT ("mean1") for blocks 3 and 6 (one score using trials from both blocks), a mean RT ("mean2") for blocks 4 and 7 (one score using trials from both blocks), and the SD of RTs in blocks 3, 4, 6 and 7 ("SD"). To calculate D: D = (mean2 - mean1)/SD. Blocks 1, 2, and 5 are practice blocks and must be discarded. 

```{r}
# Exclude blocks other than 3, 4, 6 and 7
iat_without_trials <- data_raw_iat[data_raw_iat$block_number %in% c(3, 4, 6, 7), ]

iat_without_trials$trial_reaction_time_in_ms <- as.numeric(as.character(iat_without_trials$trial_reaction_time_in_ms))

# Group
iat_complete_grouped <- iat_without_trials %>%
  group_by(unique_id) 

# Calculate mean RT: Blocks 3 and 6
iat_complete_grouped_mean <- iat_complete_grouped %>%
  filter(block_number %in% c(3, 6)) %>%
  summarise(mean1 = mean(trial_reaction_time_in_ms, na.rm = TRUE))

# Calculate mean RT: Blocks 4 and 7
iat_complete_grouped_mean<- iat_complete_grouped %>%
  filter(block_number %in% c(4, 7)) %>%
  summarise(mean2 = mean(trial_reaction_time_in_ms, na.rm = TRUE)) %>%
  full_join(iat_complete_grouped_mean, by = "unique_id")

# Calculate SD
iat_complete_grouped_sd <- iat_complete_grouped %>%
  filter(block_number %in% c(3, 4, 6, 7)) %>%
  summarise(SD = sd(trial_reaction_time_in_ms, na.rm = TRUE))

# Calculate D
data_frame_D_score <- iat_complete_grouped_mean %>%
  full_join(iat_complete_grouped_sd, by = "unique_id") %>%
  mutate(D_Score = (mean2 - mean1) / SD) %>%
  select(unique_id, D_Score) %>% 
  ungroup()

print(data_frame_D_score)
```

### Sanity check for D-Score

Check that all D scores are in the range -2 to +2. If not, revise your implementation of the scoring code. 

```{r}
if(all(data_frame_D_score$D_Score >= -2 & data_frame_D_score$D_Score <= 2)) {
  print("Sanity check passed: All D scores are in the range -2 to +2.")
} else {
  print("Sanity check failed: Revise your implementation of the scoring code.")
}
```

## Exclusion variable

### Exclusion variable for participants with incomplete trail level IAT

Create an exclusion variable and set participants with incomplete trial level IAT data to "exclude". Specifically, IAT should have 120 trials on the critical test blocks (i.e., blocks 3, 4, 6 and 7). Trials on the other (practice) blocks can be discarded.

```{r}
exclude_iat_incomplete <- iat_without_trials %>%
  group_by(unique_id) %>%
  filter(all(c(3, 4, 6, 7) %in% block_number) & n() != 120) |> 
  pull(unique_id) |> 
  unique()
```

### Exclusion variable for IAT performance 

Create an exclusion variable for IAT performance: set participants with >10% of the participants trials are < 300ms, or if their accuracy is < than 75%. Only use trials from the critical test blocks when computing these (i.e., blocks 3, 4, 6 and 7).

```{r}
exclude_iat_performance <- iat_without_trials %>%
  group_by(unique_id) %>%
  summarise(
    total_trials = n(),
    trials_below_threshold = sum(trial_reaction_time_in_ms < 300),
    accuracy = sum(trial_accuracy == "correct") / total_trials
  )
head(exclude_iat_performance)
  
unique_ids_in_data_D_score <- data_frame_D_score$unique_id

# Identify participants to exclude based on low accuracy
participants_to_exclude_low_accuracy <- exclude_iat_performance$unique_id[
  exclude_iat_performance$accuracy < 0.75 &
  exclude_iat_performance$unique_id %in% unique_ids_in_data_D_score
]

# Identify participants to exclude based on short reaction times
participants_to_exclude_short_rt <- exclude_iat_performance$unique_id[
  exclude_iat_performance$trials_below_threshold > 0.1 * exclude_iat_performance$total_trials &
  exclude_iat_performance$unique_id %in% unique_ids_in_data_D_score
]

iat_exclusion <- c(exclude_iat_incomplete, participants_to_exclude_short_rt, participants_to_exclude_low_accuracy)

# Create exclusion variable in the D_score dataset
data_iat_D_exclusions <- data_frame_D_score %>%
  mutate(exclude_iat = case_when(unique_id %in% iat_exclusion ~ "exclude", TRUE ~ "include"))
```

## Combine datasets

Combine the demographics, BFI, and IAT data into one data frame. This data frame should be one-row-one-participant. Both the mean scored and item level BFI data should be present in the dataset.

```{r}
data_processed_before_exclusions <- dat_age_gender %>%
  full_join(data_bfi_completeness, by = "unique_id") %>%
  full_join(data_iat_D_exclusions, by = "unique_id")

```

# Master exclusion variable

Create a master exclude variable from the individual exclude variables. 

```{r}
data_processed <- data_processed_before_exclusions %>%
  mutate(
    exclude_participant = case_when(
      exclude_iat == "exclude" ~ "exclude",
      exclude_impossible_data == "exclude" ~ "exclude",
      exclude_incomplete_data == "exclude" ~ "exclude",
      TRUE ~ "include"
    )
  ) %>%
  select(-c(exclude_impossible_data, exclude_iat, exclude_incomplete_data))

```

# Write to disk

Save the processed data to the data/processed/ folder as "data_processed.csv". 

```{r}
# in case this dir doesn't exist, create it
dir.create("../data/processed/")

# save data to disk in that dir
write_csv(data_processed, "../data/processed/data_processed.csv")
```

# Create codebook template for the processed data

Create a codebook for the processed data file that explains what each variable represents.

```{r}
if(!file.exists("../data/processed/data_processed_codebook.xlsx")){
  # convert the column names to a df
  codebook_template <- data.frame(variable = colnames(data_processed)) |>
    mutate(explanation = NA)
  # write to disk as an excel file
  write.xlsx(codebook_template, file = "../data/processed/data_processed_codebook.xlsx")
}
```

# Session info

```{r}
sessionInfo()
```
